---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# spectralard

<!-- badges: start -->
<!-- badges: end -->

The goal of `spectralard` is to provide efficient spectral estimation of low-rank network models from Aggregated Relational Data (ARD). 

## Installation

You can install the development version of `spectralard` from [GitHub](https://github.com/alexpghayes/spectralard) with:

```r
# install.packages("devtools")
devtools::install_github("alexpghayes/spectralard")
```

## Features

- **Efficient Spectral Estimation**: Estimate latent positions and community structures directly from ARD using `estimate_spectrum()`.
- **Sparse Graph Sampling**: Generate representative graphs from estimated models using `sample_graph()`, with support for large networks through memory-efficient chunked sampling.
- **Model-Based Simulation**: Generate synthetic ARD datasets using `simulate_ard_data()`. This function generates a network from a Degree-Corrected Stochastic Block Model (DCSBM) and simulates traits that are correlated with the underlying latent positions. ARD is then formed by aggregating these traits over the neighborhood of each node.

## Quick Start

This example demonstrates how to simulate ARD data, estimate the underlying network model, and analyze the results.

```{r quickstart-sim}
library(spectralard)

# 1. Simulate ARD data
# Uses a DCSBM and generates traits correlated with latent positions
sim <- simulate_ard_data(
  n = 200,              # Number of nodes
  k = 3,                 # Latent dimensions
  corr = 0.8,            # Correlation between traits and latent positions
  num_good_traits = 5,   # Informative traits
  num_bad_traits = 0     # Uninformative traits
)
```

The estimation returns a spectral decomposition of the underlying network model.

```{r quickstart-estimate}
# 2. Estimate the network model
estimate <- estimate_spectrum(sim$ard, sim$traits)

# The estimate contains the spectral components
estimate$Utilde[1:5, ]
estimate$Sigmatilde
```

We can use the estimated latent positions `Utilde` to recover community structure (e.g., via k-means clustering) or sample new graphs.

```{r quickstart-analysis}
# 3. Find blocks using k-means on Utilde
clusters <- kmeans(estimate$Utilde, centers = 3)

# 4. Evaluate clustering accuracy (Adjusted Rand Index)
# A helper to calculate ARI manually to avoid extra dependencies
calc_ari <- function(x, y) {
  tab <- table(x, y)
  sum_comb_n_ij <- sum(choose(tab, 2))
  sum_comb_a_i <- sum(choose(rowSums(tab), 2))
  sum_comb_b_j <- sum(choose(colSums(tab), 2))
  n <- sum(tab)
  comb_n_2 <- choose(n, 2)
  prod_comb <- (sum_comb_a_i * sum_comb_b_j) / comb_n_2
  (sum_comb_n_ij - prod_comb) / (0.5 * (sum_comb_a_i + sum_comb_b_j) - prod_comb)
}

calc_ari(clusters$cluster, sim$model$z)

# 5. Sample a new graph from the estimate
g <- sample_graph(
  estimate, 
  allow_self_loops = FALSE, 
  poisson_edges = FALSE
)
g
```

The Adjusted Rand Index (ARI) of 1.0 indicates that the clustering perfectly
recovers the true community structure from the noisy ARD. Even with only
$0.8$ correlation between traits and latent positions, and the presence
of uninformative traits, the spectral estimation accurately identifies
the underlying blocks.

## Related work

https://github.com/mpleung/ARD, https://github.com/ahoundetoungan/PartialNetwork, AER 2020 replication package https://www.aeaweb.org/articles?id=10.1257/aer.20170861