---
title: "Linear-in-Means Estimation with 2SLS and ARD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linear-in-Means Estimation with 2SLS and ARD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spectralard)
library(Matrix)
library(ivreg)
```

## Introduction

In this vignette, we demonstrate how to perform linear-in-means estimation of peer effects when the network adjacency matrix $G$ is unknown. Instead of $G$, we only observe Aggregated Relational Data (ARD), which we use to estimate the network's spectral structure. We then use this estimate to construct instruments for a two-stage least squares (2SLS) estimation using the `ivreg` package.

## The Linear-in-Means Model

Consider the standard linear-in-means model for an outcome $y$:

$$y = \alpha \iota + \beta G y + \gamma W + \delta G W + \eta X + \epsilon$$

where:
- $G$ is the row-normalized adjacency matrix.
- $W$ is a vector of node-level exogenous covariates.
- $X$ are the latent positions from the underlying network model.
- $Gy$ represents the endogenous peer effect.
- $GW$ represents the contextual effects.

## Simulation Setup

First, we simulate a network and ARD using the package's built-in simulation tools. We use $n = 1000$ nodes and $k = 2$ latent dimensions.

```{r simulation}
set.seed(42)
n <- 1000
k <- 2

# Simulate ARD data
sim <- simulate_ard_data(
  n = n, k = k, corr = 0.8,
  num_good_traits = 5, num_bad_traits = 0
)

A <- sim$A
traits <- sim$traits
ard <- sim$ard
X <- sim$s_pop$u %*% diag(sqrt(sim$s_pop$d)) # True latent positions

# True parameters for linear-in-means model
alpha <- 1
beta <- 0.5
gamma <- 2
delta <- 1
eta <- c(0.5, -0.5) # Coefficients for latent positions

# Generate exogenous covariate W
W <- rnorm(n)

# Row-normalize A to get G
row_sums <- rowSums(A)
row_sums[row_sums == 0] <- 1
G <- A / row_sums

# Solve for endogenous y: 
# y = (I - beta * G)^{-1} * (alpha + gamma * W + delta * G * W + X * eta + eps)
eps <- rnorm(n, sd = 0.5)
I <- Diagonal(n)
rhs <- alpha + gamma * W + delta * (G %*% W) + (as.matrix(X) %*% eta) + eps
y <- as.numeric(solve(I - beta * G, rhs))
```

## Spectral ARD Estimation

Next, we estimate the network spectral structure from the ARD. We assume that the node degrees (row sums of $A$) are known and use them to normalize our spectral estimate.

```{r estimation}
estimate <- estimate_spectrum(ard, traits)

# Extract estimated components
Utilde <- estimate$Utilde
Sigmatilde <- estimate$Sigmatilde

# Construct P_tilde = Utilde %*% Sigmatilde %*% t(Utilde)
# This serves as a plug-in for the unknown adjacency matrix A
P_tilde <- tcrossprod(Utilde %*% Sigmatilde, Utilde)

# Normalize P_tilde using KNOWN degrees to get G_hat
G_hat <- P_tilde / row_sums
```

## 2SLS Estimation

We use $\hat{G}^2 W$ and $\hat{G} X$ as instruments for the endogenous peer effect $Gy$. We also include $W$, $GW$, and the latent positions $X$ as covariates.

```{r 2sls}
# Prepare peer effects and instruments
Gy <- as.numeric(G %*% y)
Gw <- as.numeric(G %*% W)
G2w_hat <- as.numeric(G_hat %*% (G_hat %*% W))

# GX latent positions as instruments
Gx_hat <- as.matrix(G_hat %*% X)
colnames(Gx_hat) <- c("Gx1", "Gx2")

# Latent positions as covariates
X1 <- X[, 1]
X2 <- X[, 2]

# Perform 2SLS using ivreg
# Model: y ~ Gy + W + Gw + X1 + X2
# Instruments for Gy: G2w_hat + Gx1 + Gx2
model_2sls <- ivreg(
  y ~ Gy + W + Gw + X1 + X2 | 
    G2w_hat + Gx_hat + W + Gw + X1 + X2
)

summary(model_2sls)
```

## Conclusion

By substituting the spectral estimate $\hat{G}$ for the unknown row-normalized adjacency matrix $G$, and leveraging known degrees, we can successfully recover peer effect estimates. Including latent positions and their estimated neighborhood averages as instruments significantly aids identification.
